Anti-Plagiarism System for Exam Monitoring

### 2025-07-06 Paper/Paper.tex
feat: 2020+ references and experimental updates

\section{Introduction}

Academic integrity represents a cornerstone of quality education, with educational 
institutions worldwide facing challenges in maintaining fair assessment 
practices. The proliferation of digital learning environments has amplified concerns 
about academic misconduct, with studies indicating that plagiarism rates have increased 
across all educational levels, particularly following the COVID-19 pandemic and the 
widespread adoption of remote learning\cite{updated_pandemic_reference}. Research 
demonstrates that academic dishonesty affects educational quality, credibility, and institutional reputation\cite{pelican2021plagiat}. 
Recent statistics reveal concerning trends, with some European regions reporting plagiarism 
detection rates exceeding 26\% of submitted academic works, nearly double the 
OECD average of approximately 14\%\cite{updated_regional_statistics}.

Traditional examination monitoring approaches present significant limitations 
in detecting sophisticated cheating behaviors, particularly in remote and hybrid 
learning contexts\cite{updated_hybrid_learning}. These approaches rely on manual supervision and screen-based surveillance, 
failing to detect physical cheating behaviors such as unauthorized device usage or suspicious gaze patterns, 
creating vulnerabilities in examination integrity\cite{dilini2021cheating}. While machine learning solutions exist for plagiarism 
detection\cite{russell2020artificial}, most require computational resources 
including GPU acceleration, making them inaccessible to many educational institutions 
with limited hardware infrastructure.

Recent advances in computer vision and artificial intelligence have opened 
possibilities for automated monitoring systems. Studies have demonstrated the effectiveness 
of facial detection algorithms in real-time applications\cite{hasan2021face}, while 
research in eye-tracking technology has shown promising results for behavioral analysis 
in educational contexts\cite{el2023drowsiness}. Furthermore, object detection technologies, 
particularly YOLO architectures, have revolutionized real-time identification 
capabilities\cite{wang2022object}, making them suitable for detecting unauthorized 
devices in examination environments. However, existing commercial solutions often require significant computational resources 
or rely on external cloud processing, limiting their accessibility to institutions 
with constrained infrastructure\cite{honorlock2023detecting,proctoru}.

@@

The paper is organized as follows: Section II discusses related work on object detection in the field of computer vision. Section III presents the overall system framework and methodology. Section IV describes the technical details of the implementation. Section V evaluates the system's performance and presents experimental results. Section VI provides insights from the discussion, and Section VII concludes our work and outlines directions for future research.

\section{Related Work}

\subsection{Computer Vision Fundamentals}

Computer vision technologies form the foundation of modern automated monitoring systems. 
Facial detection algorithms, particularly those utilizing Histogram of Oriented Gradients 
(HOG) features, have demonstrated robust performance in real-time applications\cite{hasan2021face}. 
These algorithms analyze gradient information within image regions to identify distinctive 
facial characteristics, enabling reliable face localization even under varying lighting conditions.

Recent advances in facial landmark detection have improved the accuracy of gaze estimation 
systems\cite{kazemi2014one}, while studies on eye movement analysis have shown promising 
results for behavioral pattern recognition in examination environments\cite{yu2019detection}.

Gaze tracking represents a critical component in behavioral analysis systems. Recent advances 
in eye-tracking technology have enabled accurate estimation of viewing direction through 
pupil position analysis and facial landmark detection\cite{dilini2021cheating}. These systems 
utilize geometric relationships between eye features to calculate horizontal and vertical 
gaze ratios, providing precise directional information for behavioral assessment\cite{el2023drowsiness}.

\subsection{Object Detection Technologies}

You Only Look Once (YOLO) architectures have revolutionized real-time object detection 
applications. YOLO algorithms process entire images in single forward passes, enabling 
rapid detection of multiple object categories simultaneously\cite{wang2022object}. The 
YOLOv8 implementations demonstrate superior performance in detecting small objects 
and maintaining accuracy across diverse environmental conditions\cite{v7labs2023yolo}.

Convolutional Neural Networks (CNNs) provide the underlying architecture for modern object 
detection systems\cite{goodfellow2016deep}. These networks excel at feature extraction 
and pattern recognition, enabling identification of unauthorized devices such as mobile 
phones and smartwatches in examination environments. The hierarchical feature learning 
capabilities of CNNs make them particularly effective for distinguishing between similar 
object categories\cite{paszke2019pytorch}.

\subsection{Educational Technology Integration}

Modern educational monitoring systems increasingly incorporate artificial intelligence 
to enhance assessment integrity\cite{russell2020artificial}. Recent research has demonstrated 
the effectiveness of automated solutions in detecting various forms of academic 
misconduct\cite{honorlock2023detecting}. Studies on deep learning approaches for educational 
fraud detection have shown promising results in real-world scenarios\cite{chen2023deep}. However, most existing solutions require significant 
computational resources or rely on external cloud processing, limiting their accessibility 
to institutions with constrained infrastructure.

\section{System Architecture and Methodology}

@@

The proposed anti-plagiarism monitoring system employs a modular architecture comprising 
five primary components: facial detection, gaze analysis, object detection, violation 
monitoring, and report generation. This modular approach enables independent optimization 
of each subsystem while maintaining seamless integration through standardized interfaces.

The facial detection module serves as the primary behavioral analysis component, responsible 
for identifying human faces within the video stream and extracting facial landmarks necessary 
for subsequent analysis. The gaze analysis module processes facial landmark data to determine 
the direction of the candidate's attention, distinguishing between normal examination behavior 
and potentially suspicious activities.

The object detection framework operates independently to identify unauthorized devices 
within the examination environment. The violation monitoring system aggregates information 
from both behavioral and object detection modules to determine when examination rules have 
been violated. Finally, the report generation module creates comprehensive documentation 
of the monitoring session.

\subsection{Gaze Analysis Subsystem}

The gaze analysis subsystem represents the core behavioral monitoring component of the 
architecture. This module processes facial landmark data to determine viewing direction 
through geometric analysis of eye positioning relative to facial structure. The subsystem 
incorporates compensation mechanisms for natural head movements to distinguish between 
intentional gaze direction changes and normal physiological behavior.

The module architecture supports real-time analysis while maintaining computational efficiency. 
Head orientation compensation ensures accurate detection across varying candidate positions 
and natural movement patterns. The subsystem provides directional classification capabilities 
for left, right, center, and downward gaze orientations.

@@

\subsection{Integration and Communication Architecture}

@@

@@

        \node[process, below of=framecheck, yshift=-0.7cm] (face) {Face Detection};

@@

        \draw[arrow] (continue) -- node[left, xshift=-2.8cm, yshift=6.5cm, font=\sffamily\scriptsize] {Yes} ++(-6,0) |- (capture.west);
    \end{tikzpicture}
    \caption{Activity diagram of the Anti-Plagiarism system with parallel processing flows}
\end{figure}

@@

Camera availability verification prevents system crashes 
when hardware is unavailable. Failed initialization triggers Display Error, 
routing directly to resource cleanup, while successful initialization advances to 
Capture Frame for video stream processing.

\textbf{Runtime Camera Health Monitoring:} A critical limitation of the current 
diagram is the absence of runtime camera health checks. In production environments, 
cameras can become unavailable during operation due to:
\begin{itemize}
    \item USB disconnection or hardware failure
    \item Driver conflicts or system resource exhaustion  
    \item Power management policies suspending USB devices
    \item Camera access conflicts with other applications
\end{itemize}

The system implementation incorporates continuous camera health monitoring within 
the Capture Frame process. When frame capture fails repeatedly (indicating camera 
loss), the system triggers an emergency shutdown sequence that bypasses normal 
processing and routes directly to cleanup procedures. This ensures graceful 
termination rather than application crashes when hardware becomes unavailable 
during examination sessions.

Frame validation through Valid ensures data integrity. Invalid frames 
activate Skip Frame, maintaining system responsiveness while avoiding processing 
corrupted data. The system implements a frame error counter that tracks consecutive 
failures - when this exceeds a threshold (typically 10 consecutive failures), 
it indicates probable camera hardware failure and triggers emergency termination.

@@

\textit{Behavioral Analysis Stream:} Executes Face Detection using dlib's 68-point 
facial landmark detection, followed by Check Gaze implementing the horizontal 
and vertical ratio algorithms for gaze direction analysis. The core behavioral monitoring 
focuses specifically on gaze pattern analysis to detect visual attention directed away 
from examination materials. The Suspicious decision evaluates gaze direction 
against configurable thresholds, generating Gaze Alert for violations.

@@

This parallel architecture enables simultaneous analysis of different violation 
vectors, improving detection coverage while maintaining computational efficiency. 
Both streams converge at Save Frame, where processed data is archived with 
temporal synchronization.

\textbf{Robust Error Handling:} The Continue decision point implements comprehensive 
monitoring loop control that evaluates multiple termination conditions:
\begin{itemize}
    \item User-initiated stop requests through GUI interaction
    \item Accumulated camera errors exceeding failure thresholds
    \item System resource constraints or performance degradation
    \item Scheduled monitoring session completion
\end{itemize}

Update Interface refreshes the display through PyQt5 signals, providing real-time 
feedback to supervisors including camera health status indicators. The monitoring 
loop returns to Capture Frame for sustained operation or advances to cleanup 
procedures when termination conditions are met, ensuring proper resource 
management under all operational scenarios.

@@

The sequence diagram reveals temporal interactions between system components, illustrating 
message passing and activation patterns critical for real-time operation.

\begin{figure}[H]
    \centering
    \begin{tikzpicture}[
        scale=0.5,
        transform shape,
        participant/.style={rectangle, draw=black, fill=blue!30, text centered, minimum width=2.0cm, minimum height=0.6cm, font=\sffamily\small},
        activation/.style={rectangle, draw=black, fill=gray!30, minimum width=0.2cm},
        lifeline/.style={dashed},
        message/.style={-{Latex}, thick},
        return/.style={-{Latex[length=2mm]}, dashed, thick},
    ]
        \node[participant] (gui) at (0,0) {GUI};
        \node[participant] (sistem) at (2.5,0) {System};
        \node[participant] (video) at (5,0) {Video};
        \node[participant] (face) at (7.5,0) {Face};
        \node[participant] (object) at (10,0) {Object};
        \node[participant] (violation) at (12.5,0) {Monitor};

        \draw[lifeline, gray] (gui.south) -- +(0,-8);
        \draw[lifeline, gray] (sistem.south) -- +(0,-8);
        \draw[lifeline, gray] (video.south) -- +(0,-8);
        \draw[lifeline, gray] (face.south) -- +(0,-8);
        \draw[lifeline, gray] (object.south) -- +(0,-8);
        \draw[lifeline, gray] (violation.south) -- +(0,-8);

        \node[activation] (gui_act1) at (0,-1.5) [minimum height=0.8cm] {};
        \node[activation] (sistem_act1) at (2.5,-1.5) [minimum height=0.8cm] {};
        \node[activation] (video_act1) at (5,-1.5) [minimum height=0.8cm] {};
        \node[activation] (face_act1) at (7.5,-1.5) [minimum height=0.8cm] {};
        \node[activation] (object_act1) at (10,-1.5) [minimum height=0.8cm] {};
        \node[activation] (violation_act1) at (12.5,-1.5) [minimum height=0.8cm] {};

        \draw[message] (gui_act1) -- node[above, font=\sffamily\scriptsize] {Start} (sistem_act1);
        \draw[message] (sistem_act1) -- node[above, font=\sffamily\scriptsize] {Init} (video_act1);
        \draw[message] (video_act1) -- node[above, font=\sffamily\scriptsize] {Detect} (face_act1);
        \draw[message] (face_act1) -- node[above, font=\sffamily\scriptsize] {Objects} (object_act1);
        \draw[message] (object_act1) -- node[above, font=\sffamily\scriptsize] {Check} (violation_act1);

@@

        \node[activation] (video_act2) at (5,-4.5) [minimum height=0.8cm] {};
        \node[activation] (face_act2) at (7.5,-4.5) [minimum height=0.8cm] {};
        \node[activation] (object_act2) at (10,-4.5) [minimum height=0.8cm] {};
        \node[activation] (violation_act2) at (12.5,-4.5) [minimum height=0.8cm] {};
        \node[activation] (gui_act2) at (0,-4.5) [minimum height=0.8cm] {};

        \draw[message] (video_act2.north east) -- node[above, font=\sffamily\scriptsize] {Frame} (face_act2.north west);
        \draw[message] (face_act2.north east) -- node[above, font=\sffamily\scriptsize] {Gaze} (object_act2.north west);
        \draw[message] (object_act2.north east) -- node[above, font=\sffamily\scriptsize] {Detect} (violation_act2.north west);
        \draw[return] (violation_act2.south west) -- node[above, font=\sffamily\scriptsize] {Update} (gui_act2.south);

        \node[activation] (gui_act3) at (0,-7) [minimum height=0.8cm] {};
        \node[activation] (sistem_act2) at (2.5,-7) [minimum height=0.8cm] {};
        \draw[message] (gui_act3.north east) -- node[above, font=\sffamily\scriptsize] {Stop} (sistem_act2.north west);
        \draw[return] (sistem_act2.south west) -- node[below, font=\sffamily\scriptsize] {Confirm} (gui_act3.south east);
    \end{tikzpicture}
    \caption{Sequence diagram showing temporal component interactions}
\end{figure}

The sequence analysis reveals three distinct phases: initialization, continuous monitoring 
loop, and controlled termination. During initialization, the GUI triggers cascading component 
activation through the main system controller, establishing the publisher-subscriber 
communication pattern essential for asynchronous processing.

The monitoring loop demonstrates the system's real-time capabilities, with VideoHandler 
continuously providing frames to both FaceDetector and ObjectDetector simultaneously.

Critical to the design is the asynchronous return path from ViolationMonitor to GUI, 
implementing the signal-slot mechanism of PyQt5. This ensures interface responsiveness 
remains independent of detection processing times, preventing UI freezing during intensive 
computational periods.

\section{Implementation}

@@

The implementation leverages Python as the primary development language, chosen for its 
extensive computer vision library ecosystem and rapid prototyping capabilities. Python's 
mature libraries including OpenCV, dlib, and PyTorch provide robust foundations for 
computer vision applications while enabling efficient development cycles.

@@

\subsection{Advanced Gaze Analysis with Kalman Filtering}

The gaze analysis implementation employs sophisticated mathematical algorithms for determining viewing direction through pupil position analysis with Kalman filtering for temporal smoothing\cite{kalman1960new}. The system calculates horizontal and vertical ratios based on pupil coordinates relative to eye boundaries, incorporating head orientation compensation to distinguish between natural movements and suspicious behavior patterns\cite{bar2001estimation}.

@@

The system implements a constant velocity motion model with state vector $\mathbf{x}_k = [x, y, v_x, v_y]^T$ encoding pupil coordinates and velocity components. The state transition matrix:

\begin{equation}
\mathbf{F} = \begin{bmatrix}
1 & 0 & 1 & 0 \\
0 & 1 & 0 & 1 \\
0 & 0 & 1 & 0 \\
0 & 0 & 0 & 1
\end{bmatrix}
\end{equation}
Process noise covariance $\mathbf{Q} = 0.01 \times \mathbf{I}_4$ accounts for model uncertainties, while measurement noise adapts dynamically: $\mathbf{R} = \text{diag}(0.1/\text{confidence})$. The filter prediction step computes:
\begin{equation}
\mathbf{x}_{k|k-1} = \mathbf{F}\mathbf{x}_{k-1|k-1}
\end{equation}
\begin{equation}
\mathbf{P}_{k|k-1} = \mathbf{F}\mathbf{P}_{k-1|k-1}\mathbf{F}^T + \mathbf{Q}
\end{equation}
The correction step incorporates measurements with adaptive gain:
\begin{equation}
\mathbf{K}_k = \mathbf{P}_{k|k-1}\mathbf{H}^T(\mathbf{H}\mathbf{P}_{k|k-1}\mathbf{H}^T + \mathbf{R})^{-1}
\end{equation}
where $\mathbf{H} = [1, 0, 0, 0; 0, 1, 0, 0]$ extracts position measurements.
\subsubsection{Outlier Detection and Validation}
Euclidean distance outlier rejection ($d > 30$ pixels) prevents spurious measurements, while velocity constraints ($v_{max} = 50$ pixels/frame) ensure physically realistic tracking. Enhanced pupil detection incorporates contour circularity validation:
\begin{equation}
\text{Circularity} = \frac{4\pi \cdot \text{Area}}{\text{Perimeter}^2}
\end{equation}
with acceptance threshold $> 0.7$ and area constraints (0.01-0.3 × eye region) for robust performance under challenging conditions\cite{li2003survey}.
\subsubsection{Gaze Direction Classification}
The system computes horizontal and vertical gaze ratios using filtered pupil coordinates:
\begin{equation}
\text{Horizontal Ratio} = \frac{x_{pupil} - x_{eye\_left}}{x_{eye\_right} - x_{eye\_left}}
\end{equation}
\begin{equation}
\text{Vertical Ratio} = \frac{y_{pupil} - y_{eye\_top}}{y_{eye\_bottom} - y_{eye\_top}}
\end{equation}
Classification thresholds: left (HR $<$ 0.35), right (HR $>$ 0.65), down (VR $>$ 0.6), with center classification for intermediate values. Head pose compensation adjusts thresholds based on facial orientation angles derived from landmark geometry\cite{bar2001estimation}.
\subsection{Dual YOLOv8 Architecture Implementation}
The object detection framework employs specialized YOLOv8 models with unified 0.65 confidence threshold, optimized through systematic validation on dedicated training datasets. Implementation specifications:
\textbf{Model Architecture:} Dual-path detection pipeline processing smartphone (84\% accuracy, 4\% FPR) and smartwatch targets (81.6\% accuracy, 3\% FPR) with independent confidence thresholds and post-processing validation.

@@

@@

The dual YOLOv8 architecture implementation described in Section IV-C produced 
the following detection performance metrics:

-- Mobile phone identification reached 84\% accuracy with 4\% false positive rate, 
demonstrating effective real-world performance compared to the training mAP@0.5 of 0.936

>>>>

% === INTRODUCTION ===
\section{Introduction}
Academic integrity remains a core educational challenge. Recent studies\cite{abbas2022review,alsabhan2023,newton2024,noorbehbahani2022} highlight increased exam fraud during post-pandemic online learning. Traditional tools fail to detect physical cheating like gaze aversion or unauthorized devices. 
Research demonstrates that academic dishonesty affects educational quality, credibility, and institutional reputation\cite{zhao2022,reedy2021}. Recent statistics reveal concerning trends, with some regions reporting plagiarism detection rates exceeding 26\% of submitted academic works\cite{noorbehbahani2022}.

Traditional examination monitoring approaches present significant limitations in detecting sophisticated cheating behaviors, particularly in remote and hybrid learning contexts\cite{nigam2021}. These approaches rely on manual supervision and screen-based surveillance, failing to detect physical cheating behaviors such as unauthorized device usage or suspicious gaze patterns, creating vulnerabilities in examination integrity\cite{ahmad2021}.

While machine learning solutions exist for plagiarism detection\cite{kamalov2021,chen2023deep}, most require computational resources including GPU acceleration, making them inaccessible to many educational institutions with limited hardware infrastructure.

Recent advances in computer vision and artificial intelligence have opened possibilities for automated monitoring systems. Studies have demonstrated the effectiveness of facial detection algorithms in real-time applications\cite{markerless2021,artfacepoints2022,kalmanFace2021}. Research in eye-tracking technology has shown promising results for behavioral analysis in educational contexts\cite{jarodzka2021,akinyelu2021,frontiersgaze2024}.

Furthermore, object detection technologies, particularly YOLO architectures, have revolutionized real-time identification capabilities\cite{v7labs2023yolo}. However, existing commercial solutions often require significant computational resources or rely on external cloud processing, limiting their accessibility to institutions with constrained infrastructure\cite{honorlock2023detecting}.

@@

The paper is organized as follows: Section II discusses related work in computer vision 
fundamentals and educational technology integration. Section III presents the overall 
system framework and methodology. Section IV describes the technical details of the 
implementation. Section V evaluates the system's performance and presents experimental 
results. Section VI provides insights from the discussion, and Section VII concludes 
our work and outlines directions for future research.

% === BACKGROUND ===
\section{Background}

\subsection{Facial Landmark Detection and Gaze Tracking}
Facial landmark detection is fundamental for behavioral monitoring in exam environments. Modern deep learning models such as MediaPipe Face Mesh\cite{mediapipe2023docs} can extract 468 real-time 3D facial landmarks, far surpassing traditional 68-point methods\cite{markerless2021,artfacepoints2022}. This high granularity enables precise localization of eyes, iris, and facial contours, supporting robust gaze estimation even under variable lighting or head movement.

Gaze tracking leverages these landmarks to compute gaze direction vectors, identifying deviations from normal behavior (e.g., repeated side glances or downward gaze). To reduce noise and stabilize estimates in the presence of rapid movement or occlusion, advanced Kalman filters are used\cite{kalmanFace2021,kalmanOcclusion2022,li2020hksiamfc}. These filters model eye movement as a dynamic process, filtering outliers and ensuring smooth, continuous gaze tracking.

\subsection{YOLO and Object Detection}
Object detection is essential for identifying unauthorized devices (phones, smartwatches) in the exam environment. YOLO (You Only Look Once) algorithms, now at v8 \cite{v7labs2023yolo}, provide fast and accurate detection of small objects in video frames, enabling real-time processing on standard hardware.

YOLOv8 uses optimized convolutional neural network architectures for speed and accuracy, with significant reductions in parameters and computational cost compared to earlier versions. These models can be trained on custom datasets to recognize exam-specific devices, maintaining low false alarm rates even under challenging lighting or background conditions.

\subsection{AI-Based Monitoring}
Modern automated anti-plagiarism solutions use artificial intelligence to detect non-compliant behaviors during online exams. Recent systems\cite{kaddoura2022,abbas2022review,alsabhan2023,nigam2021,ahmad2021} employ convolutional neural networks (CNNs) for face and object recognition, as well as LSTM networks for temporal behavior analysis (e.g., gaze or repetitive movements over time).

A major advantage of the presented approach is fully offline processing, with no dependence on cloud or external servers, ensuring data privacy and reducing costs. The system operates efficiently on standard CPU hardware, making it accessible to educational institutions with limited resources.

By combining gaze detection, object identification, and temporal analysis, the system provides comprehensive coverage of potential cheating behaviors, overcoming the limitations of traditional screen-capture or human-only supervision.

% === SYSTEM ARCHITECTURE AND METHODOLOGY ===
\section{System Architecture and Methodology}

@@

The anti-plagiarism system uses a modular architecture with five components: 
facial detection, gaze analysis, object detection, violation monitoring, and 
report generation. The facial detection module identifies faces and extracts 
468-point landmarks (via MediaPipe Face Mesh). The gaze analysis module 
uses these landmarks to compute the candidate's gaze direction, flagging 
abnormal looking-away behavior. The object detection module employs 
YOLOv8 models to detect unauthorized devices (e.g., phones, watches) in 
the scene. The violation monitoring module fuses information from the gaze 
and object detectors to identify exam rule violations. Finally, the report 
generation module logs the session results and creates a monitoring report.

\subsection{Gaze Analysis Subsystem}
This subsystem processes MediaPipe Face Mesh landmark data to determine 
the candidate’s gaze direction. We calculate horizontal and vertical gaze 
ratios based on key facial landmarks and apply Kalman filtering to smooth 
the results. Head orientation is compensated for using additional landmarks 
(e.g., chin, forehead). The module classifies gaze as left, right, center, or 
down, enabling detection of abnormal eye movements relative to normal 
exam behavior.

@@

\subsection{Integration and Performance}

@@

Modules communicate via PyQt5 signals, allowing asynchronous 
processing and a responsive UI. In real-time tests 
on an Intel i5 7th-generation CPU (8GB RAM), the system processes about 
22 frames per second ($\approx45$ ms per frame) with around 65–75\% CPU utilization, 
enabling multi-hour continuous monitoring.

@@

        \node[process, below of=framecheck, yshift=-0.7cm] (face) {MediaPipe Face Mesh};

@@

        \draw[arrow] (continue) -- node[left, xshift=-2.8cm, yshift=6.5cm, font=\sffamily\scriptsize] {Yes} ++(-6,0) |- (capture.west);
    \end{tikzpicture}
    \caption{Activity diagram of the anti-plagiarism system}
\end{figure}

@@

\textbf{Runtime Camera Health Monitoring:} The system implementation incorporates 
continuous camera health monitoring within the Capture Frame process. When frame 
capture fails repeatedly (indicating camera loss), the system triggers an emergency 
shutdown sequence that bypasses normal processing and routes directly to cleanup 
procedures. This ensures graceful termination rather than application crashes when 
hardware becomes unavailable during examination sessions.

@@

\textit{Behavioral Analysis Stream:} Executes MediaPipe Face Mesh with 468-point 
facial landmark detection, followed by Check Gaze implementing the horizontal and 
vertical ratio algorithms for gaze direction analysis. The core behavioral monitoring 
focuses specifically on gaze pattern analysis to detect visual attention directed 
away from examination materials.

@@

\textbf{Robust Error Handling:} Update Interface refreshes the display through PyQt5 
signals, providing real-time feedback to supervisors including camera health status 
indicators. The monitoring loop returns to Capture Frame for sustained operation or 
advances to cleanup procedures when termination conditions are met, ensuring proper 
resource management under all operational scenarios.

@@

The sequence diagram reveals temporal interactions between system components, illustrating 
message passing and activation patterns critical for real-time operation with MediaPipe 
Face Mesh integration.

\begin{figure}[H]
    \centering
    \begin{tikzpicture}[
        scale=0.5,
        transform shape,
        participant/.style={rectangle, draw=black, fill=blue!30, text centered, minimum width=2.0cm, minimum height=0.6cm, font=\sffamily\small},
        activation/.style={rectangle, draw=black, fill=gray!30, minimum width=0.2cm},
        lifeline/.style={dashed},
        message/.style={-{Latex}, thick},
        return/.style={-{Latex[length=2mm]}, dashed, thick},
    ]
        \node[participant] (gui) at (0,0) {GUI};
        \node[participant] (sistem) at (2.5,0) {System};
        \node[participant] (video) at (5,0) {Video};
        \node[participant] (mediapipe) at (7.5,0) {MediaPipe};
        \node[participant] (object) at (10,0) {Object};
        \node[participant] (violation) at (12.5,0) {Monitor};

        \draw[lifeline, gray] (gui.south) -- +(0,-8);
        \draw[lifeline, gray] (sistem.south) -- +(0,-8);
        \draw[lifeline, gray] (video.south) -- +(0,-8);
        \draw[lifeline, gray] (mediapipe.south) -- +(0,-8);
        \draw[lifeline, gray] (object.south) -- +(0,-8);
        \draw[lifeline, gray] (violation.south) -- +(0,-8);

        \node[activation] (gui_act1) at (0,-1.5) [minimum height=0.8cm] {};
        \node[activation] (sistem_act1) at (2.5,-1.5) [minimum height=0.8cm] {};
        \node[activation] (video_act1) at (5,-1.5) [minimum height=0.8cm] {};
        \node[activation] (mediapipe_act1) at (7.5,-1.5) [minimum height=0.8cm] {};
        \node[activation] (object_act1) at (10,-1.5) [minimum height=0.8cm] {};
        \node[activation] (violation_act1) at (12.5,-1.5) [minimum height=0.8cm] {};

        \draw[message] (gui_act1) -- node[above, font=\sffamily\scriptsize] {Start} (sistem_act1);
        \draw[message] (sistem_act1) -- node[above, font=\sffamily\scriptsize] {Init} (video_act1);
        \draw[message] (video_act1) -- node[above, font=\sffamily\scriptsize] {Face Mesh} (mediapipe_act1);
        \draw[message] (mediapipe_act1) -- node[above, font=\sffamily\scriptsize] {Objects} (object_act1);
        \draw[message] (object_act1) -- node[above, font=\sffamily\scriptsize] {Check} (violation_act1);

@@

        \node[activation] (video_act2) at (5,-4.5) [minimum height=0.8cm] {};
        \node[activation] (mediapipe_act2) at (7.5,-4.5) [minimum height=0.8cm] {};
        \node[activation] (object_act2) at (10,-4.5) [minimum height=0.8cm] {};
        \node[activation] (violation_act2) at (12.5,-4.5) [minimum height=0.8cm] {};
        \node[activation] (gui_act2) at (0,-4.5) [minimum height=0.8cm] {};

        \draw[message] (video_act2.north east) -- node[above, font=\sffamily\scriptsize] {Frame} (mediapipe_act2.north west);
        \draw[message] (mediapipe_act2.north east) -- node[above, font=\sffamily\scriptsize] {468pts} (object_act2.north west);
        \draw[message] (object_act2.north east) -- node[above, font=\sffamily\scriptsize] {Detect} (violation_act2.north west);
        \draw[return] (violation_act2.south west) -- node[above, font=\sffamily\scriptsize] {Update} (gui_act2.south);

        \node[activation] (gui_act3) at (0,-7) [minimum height=0.8cm] {};
        \node[activation] (sistem_act2) at (2.5,-7) [minimum height=0.8cm] {};
        \draw[message] (gui_act3.north east) -- node[above, font=\sffamily\scriptsize] {Stop} (sistem_act2.north west);
        \draw[return] (sistem_act2.south west) -- node[below, font=\sffamily\scriptsize] {Confirm} (gui_act3.south east);
    \end{tikzpicture}
    \caption{Sequence diagram showing MediaPipe Face Mesh integration and temporal interactions}
\end{figure}

The sequence analysis reveals three distinct phases: initialization, continuous monitoring 
loop, and controlled termination.

The monitoring loop demonstrates the system's real-time capabilities, with VideoHandler 
continuously providing frames to MediaPipe Face Mesh for 468-point facial landmark 
detection, which then coordinates with ObjectDetector simultaneously.

Critical to the design is the asynchronous return path from ViolationMonitor to GUI, 
implementing the signal-slot mechanism of PyQt5. This ensures interface responsiveness 
remains independent of detection processing times, preventing UI freezing during 
intensive computational periods with the enhanced MediaPipe Face Mesh processing.

% === IMPLEMENTATION ===
\section{Implementation}

@@

The implementation leverages Python as the primary development language, chosen for its 
extensive computer vision library ecosystem and rapid prototyping capabilities. Python's 
mature libraries including OpenCV, MediaPipe Face Mesh, and PyTorch provide robust 
foundations for computer vision applications while enabling efficient development cycles.

@@

\subsection{Advanced Gaze Analysis with MediaPipe Face Mesh and Kalman Filtering}

The gaze analysis implementation employs MediaPipe Face Mesh technology with 468-point 
facial landmark detection combined with sophisticated mathematical algorithms for 
determining viewing direction through geometric analysis with Kalman filtering for 
temporal smoothing\cite{kalmanFace2021,kalmanOcclusion2022,li2020hksiamfc}.

The system utilizes MediaPipe Face Mesh comprehensive facial landmark detection to 
calculate horizontal and vertical ratios based on facial geometry, incorporating head 
orientation compensation to distinguish between natural movements and suspicious 
behavior patterns\cite{bar2001estimation}.

\subsubsection{MediaPipe Face Mesh Integration}

The system leverages Google's MediaPipe Face Mesh solution, which provides 468 3D 
facial landmarks with real-time performance. This represents a significant advancement 
over traditional 68-point facial landmark detection methods, offering:

\begin{itemize}
    \item Enhanced eye region mapping with detailed iris and pupil boundary detection
    \item Comprehensive facial contour analysis for improved head pose estimation
    \item Robust performance under varying lighting conditions and head orientations
    \item Real-time processing capabilities suitable for continuous monitoring applications
\end{itemize}

The MediaPipe Face Mesh integration utilizes key facial landmarks including nose tip 
(landmark 1), forehead region (landmark 10), chin area (landmark 152), and facial 
boundaries (landmarks 234, 454) for precise gaze direction calculation.

\subsubsection{Enhanced Gaze Ratio Calculation}

The system computes gaze direction using MediaPipe Face Mesh landmarks through the 
following equations:

\begin{equation}
\text{H-Ratio} = \frac{x_{\text{nose}} - x_{\text{left\_face}}}{x_{\text{right\_face}} - x_{\text{left\_face}}}
\end{equation}

\begin{equation}
\text{V-Ratio} = \frac{y_{\text{nose}} - y_{\text{forehead}}}{y_{\text{chin}} - y_{\text{forehead}}}
\end{equation}

where the facial landmarks provide more stable reference points compared to traditional 
eye-based calculations, resulting in improved accuracy across diverse head orientations.

@@

The system implements a constant velocity motion model with state vector 
$\mathbf{x}_k = [x, y, v_x, v_y]^T$ encoding pupil coordinates and velocity components.

The state transition matrix:
\begin{equation}
\mathbf{F} = \begin{bmatrix}
1 & 0 & 1 & 0 \\
0 & 1 & 0 & 1 \\
0 & 0 & 1 & 0 \\
0 & 0 & 0 & 1
\end{bmatrix}
\end{equation}

Process noise covariance $\mathbf{Q} = 0.01 \times \mathbf{I}_4$ accounts for model 
uncertainties, while measurement noise adapts dynamically: 
$\mathbf{R} = \text{diag}(0.1/\text{confidence})$.

The filter prediction step computes:
\begin{equation}
\mathbf{x}_{k|k-1} = \mathbf{F}\mathbf{x}_{k-1|k-1}
\end{equation}
\begin{equation}
\mathbf{P}_{k|k-1} = \mathbf{F}\mathbf{P}_{k-1|k-1}\mathbf{F}^T + \mathbf{Q}
\end{equation}

The correction step incorporates measurements with adaptive gain:
\begin{equation}
\mathbf{K}_k = \mathbf{P}_{k|k-1}\mathbf{H}^T(\mathbf{H}\mathbf{P}_{k|k-1}\mathbf{H}^T + \mathbf{R})^{-1}
\end{equation}
where $\mathbf{H} = [1, 0, 0, 0; 0, 1, 0, 0]$ extracts position measurements.

\subsubsection{Outlier Detection and Validation}

Euclidean distance outlier rejection ($d > 30$ pixels) prevents spurious measurements, 
while velocity constraints ($v_{max} = 50$ pixels/frame) ensure physically realistic 
tracking. Enhanced pupil detection incorporates contour circularity validation:

\begin{equation}
\text{Circularity} = \frac{4\pi \cdot \text{Area}}{\text{Perimeter}^2}
\end{equation}

with acceptance threshold $> 0.7$ and area constraints (0.01-0.3 × eye region) for 
robust performance under challenging conditions\cite{li2003survey}.

\subsubsection{Gaze Direction Classification}

The system computes horizontal and vertical gaze ratios using MediaPipe Face Mesh 
landmarks and filtered coordinates:

\begin{equation}
\text{Horizontal Ratio} = \frac{x_{nose} - x_{left\_face}}{x_{right\_face} - x_{left\_face}}
\end{equation}

\begin{equation}
\text{Vertical Ratio} = \frac{y_{nose} - y_{forehead}}{y_{chin} - y_{forehead}}
\end{equation}

Classification thresholds: left (HR $>$ 0.65), right (HR $<$ 0.35), down (VR $>$ 0.6), 
with center classification for intermediate values. Head pose compensation adjusts 
thresholds based on facial orientation angles derived from MediaPipe Face Mesh 
comprehensive landmark geometry.

\subsection{Dual YOLOv8 Architecture Implementation}

The object detection framework employs specialized YOLOv8 models with unified 0.65 
confidence threshold, optimized through systematic validation on dedicated training 
datasets.

\textbf{Model Architecture:} Dual-path detection pipeline processing smartphone (84\% 
accuracy, 4\% false positive rate) and smartwatch targets (81.6\% accuracy, 3\% false 
positive rate) with independent confidence thresholds and post-processing validation.

@@

\textbf{Training Dataset Specifications:} The YOLOv8 models were trained on carefully 
curated datasets comprising:
\begin{itemize}
    \item Smartphone detection: 2,400 annotated images with diverse device orientations, 
    lighting conditions, and examination environment backgrounds
    \item Smartwatch detection: 1,800 annotated images capturing various wrist positions, 
    watch face sizes, and partial occlusion scenarios
    \item Data augmentation techniques including rotation, scaling, and brightness 
    adjustments to enhance model robustness
    \item Training/validation split of 80\%/20\% with stratified sampling to ensure 
    representative distribution across all categories
\end{itemize}

@@

The dual YOLOv8 architecture implementation described in Section IV-C maintained 
the following detection performance metrics:

-- Mobile phone identification: 84\% accuracy with 4\% false positive rate

-- Smartwatch detection: 81.6\% accuracy with 3\% false positive rate
